<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Supplementary material for}
-->

# Spit depths and depths and spits of samples collected for radiocarbon dating at Gua Mo'o hono, Sulawesi


```{r setup, echo = FALSE}
library(knitr)
opts_chunk$set(fig.path='figure/') 
```


```{r load_data, echo = FALSE, message = FALSE}
# make sure dplyr comes first
select <- dplyr::select
library(PiperetalGMHBone)
# First load the data and code libraries. 
spit_depths <- read.csv("GMH_spit_depths.csv", stringsAsFactors = FALSE)
dates_input <- read.csv("GMH_C14_dates.csv", stringsAsFactors = FALSE)
my_libraries()
```

## Chronology of the excavated deposit

In this section we calibrate radiocarbon dates with OxCal 4.2 & IntCal13. We manaully send the raw radiocarbon ages to OxCal and merge the Oxcal output back in so we can work with the depths and ages.

```{r excal_prep}
# need to have:
# Name = lab code, Date = radiocarbon age, Uncertainty = error
dates_for_oxcal <- with(dates_input, data.frame(Name = DirectAMS.code,
                                          Date = BP,
                                          Uncertainty = X1s.error.1))
dates <- dates_for_oxcal 
oxcal_formatter(dates)
# now follow instructions in oxcal_formatter.R to use OxCal website
# now load output from OxCal website and put output in data folder

dates <- read.csv("GMH_C14_dates_OxCal_output.csv", stringsAsFactors = FALSE)
# combine with raw input table, prepare by making colnames useful
colnames(dates) <- c("Name", as.character(dates[1,])[2:7])
dates$DirectAMS.code <- gsub("R_Date ", "", dates$Name)
dates[,2:7] <- sapply(dates[,2:7], as.numeric)
dates_calib <- merge(dates_input, dates, by = 'DirectAMS.code')
# don't want to see the spit column as those are an early rough guess
# not a computed result
dates_calib <- select(dates_calib, -Excavation.unit..spit.)
# tidy up a little, remove spaces
dates_calib$material <- gsub("\\s", "", dates_calib$X)

```
Here's the OxCal output:

```{r}
dates_calib
```


### Difference between shell and organic ages

Compute separate linear models for the ages obtained from the shell and the organics, verify that the models show a non-random association of age and depth, then compute the difference in intercept. 

```{r shell_organic_dates}
# by material type

# function to extract a, b, r2, and p values

lm_eqn = function(df){
    m = lm(median ~ Depth.below.surface..m., df);
      out <-data.frame(a = coef(m)[1], 
                       b = coef(m)[2], 
                      r2 = summary(m)$r.squared,
                       p =  anova(m)$'Pr(>F)'[1],
                       n = nrow(df) )
    out                 
}
# compute linear models
models_x <- plyr::dlply(dates_calib, "material", function(i) 
     lm_eqn(i))
# present in tidy data frame
lm_coeffs <- do.call("rbind", models_x)


# for organics only: age = 3030.2 * depth -342.5
# check for differences in intercept to get offset

intercepts <- round(lm_coeffs$a,2)
difference <- round(sum(abs(intercepts)), 1)

# plot
limits <- with(dates_calib, ggplot2::aes(ymax = from, ymin = to))
ggplot(dates_calib, aes(Depth.below.surface..m., median, colour = material))   + # all, by colour
  geom_point(size = 4, aes(shape = material)) +
  geom_pointrange(limits, width = 2) +
  geom_smooth(se = FALSE, method = "lm") + 
  coord_flip() + 
  theme_minimal() + 
  scale_x_reverse() + 
  xlab("meters below surface") +
  ylab("cal yrs BP") +
  theme(axis.title.x=element_text(colour="black", size = 15),  
        # increase axis title size slightly  
        axis.title.y=element_text(colour="black", angle=90, size = 15),    
        # increase axis title size slightly and rotate
        axis.text.x=element_text(colour="black", size = 15),               
        # increase size of numbers on x-axis
        axis.text.y=element_text(colour="black", size = 15))               
        # increase size of numbers on y-axis

```

Here's a table of the linear model coefficients:

```{r}
lm_coeffs
```


The difference in intercepts in the linear models of the shell dates and organic dates is `r difference`. This indicates an approximate difference of `r difference` years between paired shell and organic dates.

If we want to use a linear model to predict age from depth, using only the ages from organic samples, the equation would be:

age = `r round(lm_coeffs[1,1],2)` + `r round(lm_coeffs[1,2],2)` * depth

### Spit depths

Convert the field measurements to be relative to the datum, then convert to to be relative to the ground surface (ie. the start depths of spit one).

```{r spit_depths}
# relate every spit depth back to the datum
spit_depths[] <- sapply(spit_depths, as.numeric)
relative_to_datum <- spit_depths$datum - spit_depths[,4:13]
# take average of start levels of spit one for
# ground surface relative to datum
ground_surface <- mean(as.numeric(relative_to_datum[1,1:4]))
# now compute relative to ground surface, this should give
# values that we can read as actual depths below the surface
relative_to_ground_surface <- relative_to_datum + abs(ground_surface)
relative_to_ground_surface$spit <- spit_depths$Spit
relative_to_ground_surface

# plot
spits_m <- melt(relative_to_ground_surface, id.var = 'spit')
spits_m$start_or_end <- ifelse(grepl("end", spits_m$variable), "end", "start")
remove <- c(".end", ".start")
spits_m$spit_start_or_end <- with(spits_m, paste0(spit, "_", start_or_end)) 
spits_m$variable <- gsub(paste(remove, collapse = "|"), "", spits_m$variable)
ggplot(spits_m, aes(variable, value, group = spit_start_or_end, linetype =  start_or_end, colour = start_or_end)) +
  geom_line() +
  theme_minimal() +
  scale_y_continuous(breaks = seq(-3,0,0.25)) +
  xlab("Measurement location") +
  ylab("Depth below surface (m)")

# how deep was spit nine?
nine_start <- round(mean(filter(spits_m, spit == 9, start_or_end == "start")$value),2)
nine_end <- round(mean(filter(spits_m, spit == 9, start_or_end == "end")$value),2)

```

The measured spit depths indicate a maximum depth below the surface of about 2.6 m. The rocky nature of the base of the excavation made it difficult to measure depth accurately and it's likely that the true depth is slightly greater. The section drawing indicates a depth of 2.7 m below the surface for the lowest points around the rocks, which would have been measured by tape measure, rather than the staff used for the spit depths. So 2.6-2.7m is a reliable maximum depth value. 

The lowest pottery is in spit nine, which is between `r -nine_start` and `r -nine_end` meters below the surface.

### Assignment of dated samples to specific spits

The dated samples were taken from the south wall and their depths can be measured from the section drawing. With these depths we can find the spit depths from the SW and SE corners that will identify the spit that correlates with the dated sample.

```{r dated_sample_spits}

dates_calib$section <- "SW."
# plot spits with dated samples
ggplot(spits_m, aes(variable, value)) +
  geom_line(aes(group = spit_start_or_end, linetype =  start_or_end, colour = start_or_end)) +
 geom_text(data = dates_calib, aes(section, -Depth.below.surface..m., label = DirectAMS.code), size = 1, hjust = 1.2) + 
  theme_minimal() +
  scale_y_continuous(breaks = seq(-3,0,0.25)) +
  xlab("Measurement location") +
  ylab("Depth below surface (m)")
```

Here we can see which spits the dated sample fall into. 

```{r}
# match sample depths up with a specific spit from SW-SE data
relative_to_ground_surface$South_means <- with(relative_to_ground_surface, apply(cbind(SE..end, SW..end), 1, mean))
South_depth_means <- sort(as.vector(na.omit(apply(cbind(relative_to_ground_surface$SE..end, relative_to_ground_surface$SW..end), 1, mean))), decreasing = FALSE)
intervals <- cbind(-dates_calib$Depth.below.surface..m., findInterval(-dates_calib$Depth.below.surface..m., South_depth_means))
means <- merge(data.frame(South_depth_means), relative_to_ground_surface, by.x = 'South_depth_means', by.y = 'South_means')
means <- arrange(means, South_depth_means)
# subset the table of spit depths with the intervals that the samples come from,
# and these are the spits that correlate with the depths that the dated samples
# were taken from in the South wall
df <- (list(means[intervals[,2], ]$spit,
           dates_calib$Depth.below.surface..m.,
           dates_calib$Name, 
           dates_calib$material,
           dates_calib$BP,
           dates_calib$to, 
           dates_calib$from))
df <- data.frame(sapply(df,'[',seq(max(sapply(df,length)))))
names(df) <- c("spit", "depth below surface", "code", "material", "uncal age","to cal BP", "from cal BP")
df
```

The implication for bone point phases is that it puts the start of phase C squarely in spit 17. Otherwise it's consistent with David's calculations. 

